Staged Abstract Interpreters

### 2019-03-26 paper/40_absinterp.tex
section 4

Based on the generic interface we presented in Section~\ref{bg_lang}, we now
describe the instantiation of an abstract interpreter. To keep the presentation
simple, we use simple abstract domains, and just establish a
context-/path-/flow-insensitive, store-widened analysis in this section ---
indeed, it is coarse but simple enough to setup a foundation for the staged
abstract interpreter we will present later. In Section~\ref{cfa}, we will see
how to regain context-/path-/flow-sensitivity \todo{path/flow} for pushdown
control-flow analysis; in Section~\ref{cases_imp}, we will see how to
instantiate an interval abstract domain for numerical analysis.

\paragraph{Abstract Instantiation}

@@

To effectively reuse the code between unstaged and staged variants, we also make
the lattice structure stage polymorphic. A lattice type class parameterizes over
an element type \texttt{E} and defines five operations for \texttt{R[E]}, such
as meet, join, and ordering relation; again, \texttt{R} is the higher-kinded
type indicating the binding-time.

\begin{lstlisting}
trait Lattice[E, R[+_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

For example, here we use powersets as the abstract domain, and the unstaged
lattice of powersets can be easily implemented as follows:

\begin{lstlisting}
implicit def SetLattice[T]: Lattice[Set[T], NoRep] = 
  new Lattice[Set[T], NoRep] {
    lazy val bot: Set[T] = Set[T]()
    lazy val top: Set[T] = throw new NotImplementedError()
    def $\sqsubseteq$(l1: Set[T], l2: Set[T]): Boolean = l1 subsetOf l2
    def $\sqcup$(l1: Set[T], l2: Set[T]): Set[T] = l1 union l2
    def $\sqcap$(l1: Set[T], l2: Set[T]): Set[T] = l1 intersect l2
  }
\end{lstlisting}

The ordering relation is to ask whether one set is a subset of the other, 
join is to union the two sets, and meet is to intersect the two sets.
Accordingly, other structures used in the rest of the paper such as tuples 
(products) and maps can be lifted to lattices element-wise or point-wise.
The code for them are elided.

@@

\subsection{Fixpoint Iteration}

>>>>

After seeing both unstaged and staged concrete interpreters, now we trun our
focus on abstract interpreters under the same framework. We will first present a
lattice framework with binding-time annotations, but for the purpose of clearly
illustrating the idea, we just use simple abstract domains such as power set
lattice or point-wise map lattice. Then we show the abstract components such as
abstract environments and abstract stores.
The abstract interpreter we present in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s, it is also
context/path/flow-insensitive by our choice.  In Section \ref{cfa}, we will
instantiate a context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitive by varying the monads, which is also applicable to our unstaged or
staged abstract interpreter. \todo{numerical/imp?}

A complete lattice over set $S$ is a tuple of $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ declares the type class for lattices.
Similar to @MonadOps@ in Section \ref{generic_if}, we introduce an additional
higher-kinded type @R[_]@ to annotate the binding-times.

\begin{lstlisting}
  trait SPLattice[S, R[_]] {
    val $\top$: R[S];  val $\bot$: R[S]
    def $\sqsubseteq$(l1: R[S], l2: R[S]): R[Boolean]
    def $\sqcup$(l1: R[S], l2: R[S]): R[S]
    def $\sqcap$(l1: R[S], l2: R[S]): R[S]
  }
  trait Lattice[S] extends SPLattice[S, NoRep]
\end{lstlisting}

In this section, we simply let @R[T] = T@ (i.e., @NoRep@) to instantiate the
unstaged lattice type class. An simple instance of lattice we used is the power
set abstract domain, where the ordering is subset relation, join is the union
operation, and meet is the intersect operation. The power set lattice can be
implemented as follows:

\begin{lstlisting}
  def SetLattice[T]: Lattice[Set[T]] = new Lattice[Set[T]] {
    lazy val $\bot$: Set[T] = Set[T]()
    lazy val $\top$: Set[T] = throw new NotImplementedError("No representation for top")
    def $\sqsubseteq$(l1: Set[T], l2: Set[T]): Boolean = l1 subsetOf l2
    def $\sqcup$(l1: Set[T], l2: Set[T]): Set[T] = l1 union l2
    def $\sqcap$(l1: Set[T], l2: Set[T]): Set[T] = l1 intersect l2
  }
\end{lstlisting}

Other lattices used in the rest of the paper, such as products and maps, can be
implemented similarly, or lifted from existing lattices element-wise or point-wise.

\paragraph{Abstract Components}

@@

\paragraph{Monads for Abstract Interpretation} nondetT

\paragraph{Primitive Operations}

@@

\paragraph{Caching and Fixpoint Iteration}
